---
title: "Project Part 1"
subtitle: "Group 9"

date: "16/09/2024"
authors:
  - name: Thimoté Dupuch
    affiliation: University of Twente
    email: t.dupuch@student.utwente.nl
  - name: Joris van Lierop
    affiliation: University of Twente
    email: j.j.w.vanlierop@student.utwente.nl
  - name: Jurre van Sijpveld
    affiliation: University of Twente
    email: j.vansijpveld@student.utwente.nl

format:
    html:
        embed-resources: true
        df-print: paged
---



### Loading libraries
```{r}
#| output: false
library(dplyr)
library(forcats)
library(vtable)
library(ggplot2)
library(plotly) # Optional
library(broom)
library(lmtest)
library(mice)
```


### Loading dataset
```{r}	
dataset <- read.csv("SMARTc.csv", sep = ";") # Without missing values
```


### Re-encode the categorical variables

```{r}	

dataset <- mutate(dataset,
  EVENT = factor(EVENT),
  EVENT = fct_recode(EVENT, "no" = "0", "yes" = "1"),

  SEX = factor(SEX),
  SEX = fct_recode(SEX, "male" = "1", "female" = "2"),

  DIABETES = factor(DIABETES),
  DIABETES = fct_recode(DIABETES, "no" = "0", "yes" = "1"),

  SMOKING = factor(SMOKING),
  SMOKING = fct_recode(SMOKING, "never" = "1", "former" = "2", "current" = "3"),

  alcohol = factor(alcohol),
  alcohol = fct_recode(alcohol, "never" = "1", "former" = "2", "current" = "3"),

  CEREBRAL = factor(CEREBRAL),
  CEREBRAL = fct_recode(CEREBRAL, "no" = "0", "yes" = "1"),

  CARDIAC = factor(CARDIAC),
  CARDIAC = fct_recode(CARDIAC, "no" = "0", "yes" = "1"),

  AAA = factor(AAA),
  AAA = fct_recode(AAA, "no" = "0", "yes" = "1"),

  PERIPH = factor(PERIPH),
  PERIPH = fct_recode(PERIPH, "no" = "0", "yes" = "1"),

  albumin = factor(albumin),
  albumin = fct_recode(albumin, "no" = "1", "micro" = "2", "macro" = "3"),

  STENOSIS = factor(STENOSIS),
  STENOSIS = fct_recode(STENOSIS, "no" = "0", "yes" = "1"),

)
```

### Description of the dataset and table of variables
The dataset is about cardiovascular health. It contains two outcomes : EVENT and TEVENT, the presence of cardiovascular events and the number of days the patient is in study until the event occurs. The dataset contains many variables, some of them are categorical and some of them are numerical. It covers patient descriptives, classical risk factors, previous symptomatic atherosclerosis, and markers of atherosclerosis.


```{r}
sumtable(dataset, out = "return", add.median = TRUE)
```



## Association between variables and the outcome

```{r}
avg_event_proportion <- mean(as.numeric(dataset$EVENT == "yes"))
bar_plot <- ggplot(dataset, aes(x = SMOKING, fill = EVENT)) +
    geom_bar(position = "fill") +
    geom_hline(yintercept = avg_event_proportion, linetype = "dashed") +
    labs(
        title = "Cardiovascular Event by Smoking Status",
        x = "Smoking Status", y = "Proportion (-- : Average)",
        fill = "Cardiovascular Event"
    ) +
    coord_flip()

ggplotly(bar_plot)
```

This bar plot shows the proportion of cardiovascular events by smoking status. The dashed line represents the average proportion of cardiovascular events in the dataset. The proportion of cardiovascular events is higher for former smokers, even higher than the average.


```{r}
boxplot <- ggplot(dataset, aes(x = as.factor(EVENT), y = AGE)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "EVENT", y = "AGE", title = "Boxplot of age by event")

ggplotly(boxplot)
```



## Logistic regression model

```{r}	
fit <- glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES +
    HISTCAR2 + HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING +
    alcohol + albumin, data = dataset, family = "binomial")
tidy(fit)
```

The logistic regression model shows that the variables associated with the lowest p-values are HISTCAR2, HDL and AGE. The other variables have higher p-values, which means they are less associated with the outcome.



```{r}
anova(fit)
```



```{r}
conf_intervals <- confint(fit, trace = FALSE, test = c("LRT", "Rao"))
conf_intervals_df <- as.data.frame(conf_intervals)
conf_intervals_df
```

```{r}
conf_intervals_df$Variable <- rownames(conf_intervals_df)
colnames(conf_intervals_df) <- c("Lower", "Upper", "Variable")
conf_intervals_df <- conf_intervals_df[conf_intervals_df$Variable != "(Intercept)", ]

conf_intervals_plot <- ggplot(conf_intervals_df, aes(x = Variable, ymin = Lower, ymax = Upper)) +
    geom_errorbar(width = 0.2) +
    geom_point(aes(y = (Lower + Upper) / 2)) +
    coord_flip() +
    labs(
        title = "Confidence Intervals for Logistic Regression Coefficients",
        x = "Variable",
        y = "Confidence Interval"
    ) + 
    theme_minimal()

conf_intervals_plot
```


## Imputation of missing values

```{r}
dataset <- read.csv("SMART.csv", sep = ";") # With missing values
```

```{r}
#| output: false
dataset_imputed <- mice(dataset, m = 5, seed = 123)
dataset_imputed <- complete(dataset_imputed)
```


```{r}
fit_imputed <- with(dataset_imputed, glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES +
    HISTCAR2 + HOMOC + log(CREAT) + STENOSIS + IMT + SMOKING +
    alcohol + albumin, data = dataset, family = "binomial"))
tidy(fit_imputed)
```


Comparison between the logistic regression model with and without imputation of missing values :

```{r}
anova(fit_imputed)
```

```{r}
conf_intervals_imputed <- confint(fit_imputed, trace = FALSE, test = c("LRT", "Rao"))
conf_intervals_imputed_df <- as.data.frame(conf_intervals_imputed)
conf_intervals_imputed_df
```

```{r}
conf_intervals_imputed_df$Variable <- rownames(conf_intervals_imputed_df)
colnames(conf_intervals_imputed_df) <- c("Lower", "Upper", "Variable")
conf_intervals_imputed_df <- conf_intervals_imputed_df[conf_intervals_imputed_df$Variable != "(Intercept)", ]

conf_intervals_imputed_plot <- ggplot(conf_intervals_imputed_df, aes(x = Variable, ymin = Lower, ymax = Upper)) +
    geom_errorbar(width = 0.2) +
    geom_point(aes(y = (Lower + Upper) / 2)) +
    coord_flip() +
    labs(
        title = "Confidence Intervals for Logistic Regression Coefficients (Imputed)",
        x = "Variable",
        y = "Confidence Interval"
    ) +
    theme_minimal()

conf_intervals_imputed_plot
```

## Plot of comparison between the two models

```{r}
conf_intervals_df$Model <- "Without Imputation"
conf_intervals_imputed_df$Model <- "With Imputation"
conf_intervals_combined <- rbind(conf_intervals_df, conf_intervals_imputed_df)

conf_intervals_combined_plot <- ggplot(conf_intervals_combined, aes(x = Variable, ymin = Lower, ymax = Upper, color = Model)) +
    geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.5,alpha = 0.5) +
    geom_point(aes(y = (Lower + Upper) / 2), alpha = 0.5) +
    coord_flip() +
    labs(
        title = "Confidence Intervals for Logistic Regression Coefficients",
        x = "Variable",
        y = "Confidence Interval"
    ) + 
    theme_minimal()

conf_intervals_combined_plot
```


---

### Final part

```{r}	
# Load the dataset
Patients_C <- read.csv("SMART.csv", sep = ";")

# Categorical variable conversions
Patients_C <- Patients_C %>%
  mutate(
    SEX = factor(SEX) %>% fct_recode("male" = "1", "female" = "2"),
    DIABETES = factor(DIABETES) %>% fct_recode("No" = "0", "Yes" = "1"),
    SMOKING = factor(SMOKING) %>% fct_recode("never" = "1", "former" = "2", "current" = "3"),
    alcohol = factor(alcohol) %>% fct_recode("never" = "1", "former" = "2", "current" = "3"),
    CEREBRAL = factor(CEREBRAL) %>% fct_recode("No" = "0", "Yes" = "1"),
    CARDIAC = factor(CARDIAC) %>% fct_recode("No" = "0", "Yes" = "1"),
    AAA = factor(AAA) %>% fct_recode("No" = "0", "Yes" = "1"),
    PERIPH = factor(PERIPH) %>% fct_recode("No" = "0", "Yes" = "1"),
    albumin = factor(albumin) %>% fct_recode("no" = "1", "micro" = "2", "macro" = "3"),
    STENOSIS = factor(STENOSIS) %>% fct_recode("No" = "0", "Yes" = "1")
  )

# Final pre-processing
final_part_c <- Patients_C %>%
  mutate(log_CREAT = log(CREAT)) %>%
  select(-c(HISTCARD, CREAT))  # Exclude these columns

```

```{r}
# Impute missing values with 'mice' - generate 5 imputed datasets
imputed_data <- mice(final_part_c, m = 5, maxit = 5, method = "pmm", seed = 123)

# Fit logistic regression model on imputed datasets
log_reg_model <- with(imputed_data, glm(EVENT ~ AGE + SEX + BMI + SYSTH + HDL + DIABETES + HISTCAR2 + HOMOC + log_CREAT + STENOSIS + IMT + SMOKING + alcohol + albumin,
    family = binomial
))

# Pool the results from the 5 imputed datasets
pooled_results <- pool(log_reg_model)

# View the pooled summary
summary(pooled_results)
```