---
title: "Project Part 1"
subtitle: "Group 9"

date: "16/09/2024"
authors:
  - name: Thimoté Dupuch
    affiliation: University of Twente
    email: t.dupuch@student.utwente.nl
  - name: Joris van Lierop
    affiliation: University of Twente
    email: j.j.w.vanlierop@student.utwente.nl
  - name: Jurre van Sijpveld
    affiliation: University of Twente
    email: j.vansijpveld@student.utwente.nl

format:
    html:
        embed-resources: true
        df-print: paged
---



### Loading libraries
```{r}
#| output: false
library(dplyr)
library(forcats)
library(vtable)
library(ggplot2)
library(plotly) # Optional
```


### Loading dataset
```{r}	
dataset <- read.csv("SMARTc.csv", sep = ";") # Without missing values
```


### Re-encode the categorical variables

```{r}	

dataset <- mutate(dataset,
  EVENT = factor(EVENT),
  EVENT = fct_recode(EVENT, "no" = "0", "yes" = "1"),

  SEX = factor(SEX),
  SEX = fct_recode(SEX, "male" = "1", "female" = "2"),

  DIABETES = factor(DIABETES),
  DIABETES = fct_recode(DIABETES, "no" = "0", "yes" = "1"),

  SMOKING = factor(SMOKING),
  SMOKING = fct_recode(SMOKING, "never" = "1", "former" = "2", "current" = "3"),

  alcohol = factor(alcohol),
  alcohol = fct_recode(alcohol, "never" = "1", "former" = "2", "current" = "3"),

  CEREBRAL = factor(CEREBRAL),
  CEREBRAL = fct_recode(CEREBRAL, "no" = "0", "yes" = "1"),

  CARDIAC = factor(CARDIAC),
  CARDIAC = fct_recode(CARDIAC, "no" = "0", "yes" = "1"),

  AAA = factor(AAA),
  AAA = fct_recode(AAA, "no" = "0", "yes" = "1"),

  PERIPH = factor(PERIPH),
  PERIPH = fct_recode(PERIPH, "no" = "0", "yes" = "1"),

  albumin = factor(albumin),
  albumin = fct_recode(albumin, "no" = "1", "micro" = "2", "macro" = "3"),

  STENOSIS = factor(STENOSIS),
  STENOSIS = fct_recode(STENOSIS, "no" = "0", "yes" = "1"),

)
```

### Description of the dataset and table of variables
The dataset is about cardiovascular health. It contains two outcomes : EVENT and TEVENT, the presence of cardiovascular events and the number of days the patient is in study until the event occurs. The dataset contains many variables, some of them are categorical and some of them are numerical. It covers patient descriptives, classical risk factors, previous symptomatic atherosclerosis, and markers of atherosclerosis.


```{r}
sumtable(dataset, out = "return", add.median = TRUE)
```



## Association between variables and the outcome

```{r}
avg_event_proportion <- mean(as.numeric(dataset$EVENT == "yes"))
SMOKING <- dataset$SMOKING
EVENT <- dataset$EVENT

bar_plot <- ggplot(dataset, aes(x = SMOKING, fill = EVENT)) +
    geom_bar(position = "fill") + # "fill" scales the bars to proportion
    geom_hline(yintercept = avg_event_proportion, linetype = "dashed") +
    labs(
        title = "Cardiovascular Event by Smoking Status",
        x = "Smoking Status", y = "Proportion (-- : Average)",
        fill = "Cardiovascular Event"
    )

ggplotly(bar_plot,width = 500, height = 600)
```