---
title: Project Part 4
date: today
subtitle: "Group 9"

authors:
  - name: Thimoté Dupuch
    affiliation: University of Twente
    email: t.dupuch@student.utwente.nl
  - name: Joris van Lierop
    affiliation: University of Twente
    email: j.j.w.vanlierop@student.utwente.nl
  - name: Jurre van Sijpveld
    affiliation: University of Twente
    email: j.vansijpveld@student.utwente.nl

format:
    html:
        embed-resources: true
        df-print: paged
        monofont: monospace
        theme: zephyr
        highlight-style: github

toc: true
toc-depth: 3
toc-expand: true
toc-location: right
---

```{r}
#| output: false
library(tidyverse) # For data manipulation
library(knitr) # For tables
library(caret) # For model training
library(pROC) # For ROC analysis
library(MASS) # For stepAIC
library(boot) # For bootstrapping
library(survival) # For survival analysis
library(rms) # For validating Cox model
library(glmnet) # For lasso and ridge regression
library(data.table) # For faster CSV read
```


### Load dataset (and re-encode variables, not needed)
```{r}
Patients <- fread("SMARTc.csv", sep = ";")
final_patients <- Patients %>%
    mutate(log_CREAT = log(CREAT)) %>%
    subset(select = -c(HISTCARD, HISTCAR2, CREAT))
```


# Cox model with stepwise selection

```{r}
#| output: false
surv_object <- Surv(time = final_patients$TEVENT, event = final_patients$EVENT)
surv_fit_formula <- surv_object ~ SEX + DIABETES + SYSTH + DIASTH + WEIGHT + BMI + LDL + TRIG + log_CREAT

initial_cox_model <- coxph(surv_fit_formula, data = final_patients)
stepwise_cox_model <- stepAIC(initial_cox_model, direction = "backward")
```

```{r}
summary(stepwise_cox_model)
```


The "Coef" represents the levels of risk. With a negative coefficient, we know the variable has a decreased risk on the event, and a positive coefficient indicates an increased risk on EVENT. Subsequently, when looking at exp(coef), we conclude that a ratio bigger than 1 means an increased risk and a ration smaller than 1 means a decreased risk.
For instance, we see that diabetes has a positive coefficient 0.304, with ratio 1.355. People with diabetes thus have a 35.5% higher risk of EVENT, which is also significant with p=0.00501

Below, we see that for instance for log_CREAT, the hazard ratio is 2.722 with a 95% confidence interval of (2.224, 3.331).

The level of concordance with 0.684 says how well the model predicts. An accuracy of 68.4% is quite well. We also see that the model is overall highly significant with the very low p-values.


# Extracting coefficients from the final model

```{r}
final_coefficients <- coef(stepwise_cox_model)
print(final_coefficients)
```

The coefficients above are log-hazard ratios, and we can calculate the hazard ratio by exponentiating each coefficient:
Hazard Ratio (HR) = ecoefficient

This means per variable, that risk is increased or decreased if the predictor variable increased. Below, the changes and Hazard ratios per variable are shortly outlined.
SEX (-0.379): e^(-0.379) = 0,684. This means that a 1-unit change (male to female) lowers the risk of occurring event with around 32%.
DIABETES (0.304): e^(0.304) = 1.355. This means that having diabetes increases the hazard. Having diabetes is associated with 35.5% higher risk of the event occurring.
SYSTH (0.017): e^(0.017) = 1.017. This indicates a 1.7% risk increase for each unit increase of the systolic blood pressure.
WEIGHT (-0.009): e^(-0.009) = 0.991.  For interpretation of this, each unit of increase in weight means a 0.9% lower risk of event occurring.
LDL (0.103): e^(0.103) = 1.109. This tells us that there’s a 10.9% higher risk of event occurring for each unit increase n LDL cholesterol.
log_CREAT (1.002) = e^(1.002) = 2.724. This indicates a 172.4% higher risk of event occurring for each 1-unit increase of the log-transformed creatinine.
Concluding. Some factors like diabetes, LDL and creatinine significantly increase the risk of occurring when unit of measuring change, while others might reduce the risk. Understanding how these variables interact with the outcome helps guide medical or practical decisions related to the modeled event.


# Calculate concordance index for the final model

```{r}
final_surv <- Surv(time = final_patients$TEVENT, event = final_patients$EVENT)
final_surv_fit_formula <- final_surv ~ predict(stepwise_cox_model, newdata = final_patients)
final_cox_model <- coxph(final_surv_fit_formula, data = final_patients)
concordance_index <- final_cox_model$concordance
print(paste("Concordance Index:", concordance_index))
```

We see from the output 7 different concordance levels. We see that the 1st to the 5th are not possible and the 7th isn't possible either. We conclude to have a concordance level of 0.648

# Summary of the final model and exponentiated coefficients
```{r}
summary(final_cox_model)
exp(final_coefficients)
```

The main takeaways of this summary are that the concordance indext of 0.648 suggests moderate predictive ability. This means that the model can differentiate between high- and low-risk individuals about 68.4% of the time.
On top of that, the complete model is statistically significant (z-score of 13.91 and p-value of <2e^(-16)). 
Lastly, as already explained in previous steps, the hazard ratios are again mentioned indicating increases and decreases in event occurring risks for 1-unit variable changes.

# Cross-validation of the Cox model

```{r}
#| output: false
set.seed(123)
folds <- createFolds(final_patients$EVENT, k = 10, list = TRUE, returnTrain = TRUE)
cv_concordance_indices <- numeric(10)

for (i in 1:10) {
    # Split the data into training and testing sets
    train_data <- final_patients[folds[[i]], ]
    test_data <- final_patients[-folds[[i]], ]

    # Create the survival object for training data
    surv_train <- Surv(time = train_data$TEVENT, event = train_data$EVENT)
    surv_fit_formula <- surv_train ~ SEX + DIABETES + SYSTH + DIASTH + WEIGHT + BMI + LDL + TRIG + log_CREAT

    # Fit the Cox model on the training data
    cox_model <- coxph(surv_fit_formula, data = train_data)

    # Predict on the test data
    test_surv <- Surv(time = test_data$TEVENT, event = test_data$EVENT)
    predictions <- predict(cox_model, newdata = test_data)

    cv_concordance_indices[i] <- coxph(test_surv ~ predictions, data = test_data)$concordance
}

mean_concordance_cv <- mean(cv_concordance_indices)
mean_concordance_cv
```

```{r}
#| output: false
bootstrap_cox_model <- function(data, indices) {
    resampled_data <- data[indices, ]
    surv_object <- Surv(time = resampled_data$TEVENT, event = resampled_data$EVENT)
    surv_fit_formula <- surv_object ~ SEX + DIABETES + SYSTH + DIASTH + WEIGHT + BMI + LDL + TRIG + log_CREAT

    initial_cox_model <- coxph(surv_fit_formula, data = resampled_data)
    stepwise_model <- stepAIC(initial_cox_model, direction = "backward", trace = FALSE)
    all_vars <- colnames(data)[-c(1, 2)]
    selected_vars <- names(coef(stepwise_model))

    # Create a binary vector: 1 if the variable is included, 0 otherwise
    inclusion_vector <- as.integer(all_vars %in% selected_vars)

    return(inclusion_vector)
}

# Set the number of bootstrap samples
set.seed(123)
n_bootstrap_samples <- 100

# Run bootstrapping
bootstrap_results <- boot(data = final_patients, statistic = bootstrap_cox_model, R = n_bootstrap_samples)
```


# Analyze results: Create a table of inclusion frequencies

```{r}
inclusion_frequencies <- colMeans(bootstrap_results$t) # Get inclusion proportions
inclusion_variables <- colnames(final_patients)[-c(1, 2)]
inclusion_frequencies <- data.frame(Variable = inclusion_variables, Proportion = inclusion_frequencies)
inclusion_frequencies
```


# Optional: Visualize the inclusion frequencies

```{r}
ggplot(inclusion_frequencies, aes(x = reorder(Variable, -Proportion), y = Proportion)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    labs(
        title = "Variable Inclusion Frequencies in Bootstrapped Cox Models",
        x = "Predictor Variables",
        y = "Proportion of Inclusion"
    ) +
    theme_minimal()
```

This code and visualisations above involves bootstrapping the Cox proportional hazards model multiple times, and tracks how frequently each variable is selected (i.e., included) in the model across different bootstrap iterations. The variable inclusion frequency plot shows how often each predictor was selected across multiple bootstrapped Cox models. It can be seen that variables like SYSTH, log_CREAT, and SEX were included more frequently, indicating they have a stronger or more consistent effect on the outcome.